Application of Micro-Rotors to Induce Crystal Nucleation: Challenges, Takeaways, and Future Prospects\\ PhD Thesis

### 2025-06-27 Latex_files/Thesis/chapters/chapter 5/main.tex
27/06/2025 chapters 1,3,4 and structural changes

\chapter{Detection of rotational spherical aggregate rotational dynamics}
\label{chapter:simulated_detection}
As outlined in the end of Chapter 4, one of the difficulties 
in characterising interactions with asymmetric objects is 
the coupled motion between translation and rotation. In order
to characterise the optical trap a position detection system 
such as a lateral effect photodiode or a quadrant photo diode
(see fig.\ref{fig:position_detectors}). Typically, a position 
detection system assumes that there exists a linear relationship 
between the a particle's displacement and the detected signal.
However, this is not always the case as demonstrated in chapter
4; as dimer's demonstrate non-harmonic trapping beyond a certain 
displacement. In addition, the rotational effects of the trapped
dimer become a more significant factor when characterising the 
trap strength. This can have unintended effects when it comes to
areas of research that require a precise understanding of the 
hydrodynamic behaviour of particles. As such there has been a 
concerted effort to either eliminate rotational motion, or describe
how the rotational behaviour influences the trajectory reported 
by the position detector.

In this chapter we build upon the work of \cite{Rohrbach2002} and
\cite{BarZiv1997}; the former used to measure the relative position
of the trapped aggregate, and the latter to measure the orientation
of the aggregate. This was initially developed in tandem with the 
work conducted in chapter 3, with the idea being that if a crystal 
nucleus formed on a rotating sphere how would that nucleus influence 
the motion of the rotating sphere. Using static light scattering we 
can map the outputted signal from a optically trapped dimer to its 
expected orientation in real time. Thus allowing us to estimate the 
optical torque being applied to the dimer by the trapping beam. As demonstrated in chapter 4, the influence of a circularly polarised 
beam can result in gyroscopic precession for specifically sized 
dimers.

We start by describing a quadrant photo diode system simulating the 
power spectrum derived from a particle trajectory. We demonstrate 
that even in idealised circumstances the rotational motion is not 
revealed from the collected power spectrum. We then go about 
defining an optical fibre detection system from a mathematical 
perspective. Before demonstrating how it can be used to 
instantaneously predict orientational information on a trapped 
particle. We then consider common factors of error in the 
characterisation process, such as signal error, and incorrect particle 
sizing. Both of which have a significant impact on the performance 
of the model, but by utilising Bayesian inference, and time averaging 
we can better refine our model.

@@

At just a cursory glance it would appear that there is a clear
relationship between the QPD signal and the orientation of the 
dimer. If we look at the QPD signal produced by a dimer as it is 
rotated we can see that it is equally sensitive to rotations in
either the X-Z or Y-Z plane but there is no change in the signal 
when rotated by in the X-Y plane. 
\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{rotation_figure}
	\caption{QPD signals detected by a dimer being rotated while
			centred at the focus of the optical trap. Top right:
			rotation in the X-Z plane. Top left: rotation in the
			Y-Z plane. Bottom: Rotation in the X-Y plane.}
	\label{fig:QPD_rotation}
\end{figure}

@@

\section{Monitoring Stochastic rotational motion using static 
		light scattering}
Part of the optical fibre arrangement borrows code from the 
optical tweezer simulations discussed in Chapter 2 and 4. 
Rather than repeat the simulation details we will instead 
only focus on the specifics of the optical fibre scattering.

\subsection{Coordinate System}
Consider a particle (either a sphere or dimer) trapped by
a Gaussian beam, with its focal point being set at [0,0,0]. 
At the same time an optical fibre directs a plane wave that 
is incident on the trapped particle scattering light in all 
directions. The probe beam is assumed to be x-polarised and 
propagating in the +z direction, this is a $90^{\circ}$ 
rotation from the previous simulations (see fig\emph{X.X}).

\subsection{Dimer}
The dimer is defined by two spheres with refractive indices
of 1.59 and suspended in water ($n_{med} = 1.33$). \textit{MSTM}
requires a target origin for setting the scattering expansion, 
there is no agreed upon solution for where the target origin 
should be relative to the sphere positions. In chapter 4 we showed
that altering the target origin would drastically alter the 
tmatrix produced. We choose (somewhat arbitrary) to use 
the dimer's centre of diffusion which is located $\zeta a_I$ from 
the centre of the larger sphere. For a discussion see \emph{X.X}.

\subsection{Scattering and Detection}
Each detector is placed some distance $\bf r$ from the origin, 
detector positions are defined by angles $\theta$ \& $\phi$. With 
$\theta=\phi=0$ being directly in front of the probe beam. The 
position of each detector is given by:
\begin{align}
	\begin{pmatrix}
		x_{fiber} \\ y_{fiber} \\ z_{fiber}
	\end{pmatrix} = 
	\begin{pmatrix}
		rcos(\phi)sin(\theta) \\ rsin(\phi)sin(\theta) \\ rcos(\theta)]
	\end{pmatrix}
\end{align}
Where r is the distance between the detector and the target origin. 
In an ideal situation the surface of the detector is oriented towards 
the target origin perfectly. Pixels can be thought of as points lying 
on the surface of the detector, each point has its own position in real
space. Using \textit{mstm} we can find the components of the scattering 
matrix at each point and compute the intensity of the electric field via:

\begin{align}
	\begin{pmatrix}
		I_s \\ Q_s \\ U_s \\ V_s
	\end{pmatrix} = 
	\begin{pmatrix}
		S_{11} & S_{12} & S_{13} & S_{14} \\
		S_{21} & S_{22} & S_{23} & S_{24} \\
		S_{31} & S_{32} & S_{33} & S_{34} \\
		S_{41} & S_{42} & S_{43} & S_{44} \\
	\end{pmatrix}
	\begin{pmatrix}
		I_i \\ Q_i \\ U_i \\ V_i
	\end{pmatrix}
\end{align}
Where:
\begin{align}
	I_s &= E_{x,\ scat}^2+E_{y,\ scat}^2  \label{eq:intensity} 
	\\
	\rightarrow I_s &= S_{11}I_i+S_{12}Q_i+S_{13}U_i+S_{14}V_i
\end{align}

By computing the scattering at each point on the detectors surface 
we can compute the average scattering over the detector. From a
experimental perspective the distance from the particle to the 
detectors must be considered. The near field scattering can be 
computed with relative ease using \textit{mstm} as shown in 
\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]{near_field_intensity_side_profile.png}
	\caption{Intensity distribution (as defined by \eqref{eq:intensity}) 
		from a symmetric dimer when irradiated by the probe beam 
		propagating in the $k\hat{z}$ direction (viewed in the 
		z-y plane). Axis are given in dimensionless units of 
		$k\hat{z}$ and $k\hat{y}$.}
	\label{fig:nf_scattering}
\end{figure}

While \textit{mstm} does provide the amplitude matrix it is not
scaled to account for distance from the target to the detector.
We amend this by simply scaling according to the inverse square 
law. As the detector is moved further from the target the 
intensity distribution across the entire detector will decrease
substantially. Because of the fall off it is more computationally
efficient to instead consider only the central pixel of the 
detector rather than across the entire surface. If we consider 
the difference between the average intensity and the intensity 
of the central point we can see that after not even $20 \mu$ 
away from the target particle the average intensity across the 
surface is within $0.1\%$ of the central pixel. 
\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]{fall_off_with_distance.png}
	\caption{Normalised difference between the central pixel intensity
	$I(\theta, \phi)$ and the average pixel intensity over the surface
	of the optical fibre $\langle I \rangle$ as a function of distance.
	4 different optical detectors were tested (spherical coordinates 
	are given in the legend). While the choice of coordinates is 
	arbitrary they all demonstrate the same exponential decay. Dashed
	lines indicate the difference after $20\mu m$ from the trap focus,
	in most practical cases getting an optical fibre any closer would
	disrupt the trap stability.}
\end{figure}

Given the trapping beam's beam waist is approximately $0.54 \mu m$,
a optical fibre would need to be a minimum of $150 \mu m$ from the 
trap centre in order to prevent noise from the trapping laser itself.
At such a distance the variation in the signal is so small that it is 
easier to evaluate the intensity at the centre of the detector instead
of across the entire surface. This is how the instantaneous orientation
is predicted later on in the chapter.

\section{Interpretation of scattering data into orientation estimates}
We now discuss the methodology for computing the instantaneous 
orientation of an optically trapped particle based on scattering
detected by a finite grouping of optical fibres.

Consider a dimer in the optical trap (Fig.~\ref{fig:dimer}a), 
we can define at any point in time a unit vector $\hat{s}$ 
pointing from the centre of the larger sphere to the centre 
of the smaller sphere. A plane wave probe beam incident, 
is incident on the dimer, generating a scattering pattern
dependent on the dimer's orientation $I(\hat{\bf s}, 
\theta, \phi)$ which is computed using \textit{mstm} 
\cite{I.Mishchenko1996}. To represent the experimental set 
up consisting of a set of optical fibres recording scattered 
light, we choose four sets of spherical angles [$(\theta_1,\ 
\phi_1), \ (\theta_2,\ \phi_2), \ (\theta_3, \ \phi_3), \ 
(\theta_4,\ \phi_4)$] and record the calculated intensity at 
each angle $I(\hat{s},\ \theta_k,\ \phi_k)$. 
\begin{figure}[h!]
	\centering
	\begin{subfigure}{0.49\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{fig2a.png}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{fig2b.png}
	\end{subfigure}
	\caption{(a)Example dimer in orientation $\hat{\bf s}$, (b) 30 Reference orientations represented by vectors pointing from [0,0,0] to each point}
	\label{fig:dimer}
\end{figure}

Our goal is to determine the orientation of the trapped dimer 
based on the measured intensity $I(\hat{n}, \ \theta_k,\ \phi_k)$. 
Rather than aim an exact estimate of the dimer's orientation, 
for the purposes of interpretation of the scattering and 
optimisation of the measurement setup it is more convenient 
to discretize the possible orientation space into a number 
of possible reference orientations. We can then use each as 
'classification categories' in a neural network methodology 
to map scattering data to orientation (see below for further 
discussion). Here we choose $\textit{n}_{ref} \ = \ 30$ 
reference orientations $\hat{\bf n}_{\alpha}$  evenly 
distributed on a unit sphere \cite{Reyuthor2006} 
(Figure~\ref{fig:dimer}b) leading to a maximum nearest 
neighbour spacing between two neighbouring reference 
orientations of 0.895 radians.

Using \textit{mstm} we compute the raw intensities (see 
\eqref{eq:intensity}) at each of the measurement angles 
that would be generated by a dimer in each reference 
orientation, $I_s(\hat{\bf n}_{\alpha}, \theta_k)$. While 
the number and position of detection fibres is technically 
arbitrary there are several constraining factors that limit 
our ability to infer useful information from the trapped 
object, see Section~\ref{sec:brownian} for a detailed 
breakdown of our choice of detection angles. In order for 
a neural network to properly weight each signal the raw 
intensities are normalized according to:
\begin{align}
	\label{eq:scale}
	y_k(\hat{\bf n}_\alpha)
	= 
	\frac{I(\hat{\bf n}_\alpha, \theta_k) - \langle I(\hat{\bf n},\theta_k) \rangle } 
	{\langle I^2(\hat{\bf n},\theta_k) \rangle -\langle I(\hat{\bf n}, \theta_k)\rangle^2}
\end{align}

where the denominator is simply the standard deviation across 
the set of values $I(\hat{\bf n}, \theta_k)$. Note that the 
collected scattering signals are not necessarily simply 
related to their associated reference orientations: as is well 
known from such examples of the inverse scattering problem. 
While it is trivial to compute the light scattering pattern 
for any given particle with any particular characteristic (i.e. 
size, shape, or orientation), inferring the light scattering 
from a unknown particle to determine said characteristic is 
incredibly difficult due to complex mapping between scattering 
and said characteristic. Even if the orientation space is 
divided evenly between reference orientations the subsequent 
signal space ends up being appearing mixed making simple 
comparisons of signals useless for inferring information on the 
particle.

Fig~\ref{fig:mixing} shows two clusters of orientation 
vectors and their respective signals collected from 3 detectors
- the points have been coloured based on their proximity 
to the centre of their respective cluster. While the 
orientation space appears tightly packed and ordered the 
signal space quickly spreads out in an asymmetric fashion. 
Furthermore as seen in Fig~\ref{fig:mixing}b the signal 
mapping can intersect itself which only further increases 
the complexity. While in some instances the mapping between 
one reference orientation and another is discrete, in other 
instances the mapping becomes far more complex to discern.
\begin{figure}[h!]
	\centering
	\begin{subfigure}{0.74\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{fig3a.png}
	\end{subfigure}
	\begin{subfigure}{0.74\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{fig3b.png}
	\end{subfigure}
	\caption{(a) Distribution of orientation vectors and (b) their respective scattering signals. Points are coloured according to their distance from the centre of each cluster (red points centred around [$0.00, 0.00, 1.00$], blue points centred at [$0.71, 0.00, 0.71$])}
	\label{fig:mixing}
\end{figure}

Nevertheless, at least where the uncertainty in signal measurements 
is low, we can predict the orientation from the scattering scattering 
by utilising computational techniques such as neural networks. We 
thus utilised the Python machine learning program \textit{scikit-learn} \cite{Pedregosa2011} to build a neural network for identifying the 
dimer's orientation from its light scattering signal. The network 
was trained by generating a database of random orientation vectors, 
calculating the corresponding light scattering signals, and then 
using the network to estimate the probability of a given signal coming 
from a dimer in a given reference orientation. The network's loss 
function was evaluated and used to improve the estimation, the 
network being trained until the improvement in the loss function 
was less than 0.0001.

Importantly, the estimation provided by the neural network can be 
improved further by accounting for any prior information we know 
about the dimer, utilising Bayesian inference to update the neural 
network's estimation: 
\begin{align}
	\label{eq:bayes}
	p(\hat{\bf n}_\alpha| y_k(\hat{\bf s}))
	&=
	\frac{p(y_k(\hat{\bf s})|\hat{\bf n}_\alpha)
		p(\hat{\bf n}_\alpha)}{p(y_k(\hat{\bf s}))}
\end{align}

where $p(\hat{\bf n}_\alpha)$ and $p(y_k(\hat{\bf s}))$ are the prior 
estimates of the distributions of particle orientations and 
instantaneous signals, respectively. \textit{Without} any prior 
evidence we must assume that the orientation prior of the dimer 
$p(\hat{\bf n}_{\alpha})$ is uniform. However, inference about the 
dimer's possible current orientation from knowledge of previous 
measurements can be used to inform our estimate of $p(\hat{\bf 
n}_{\alpha})$. The latter prior $p(y_k(\hat{s}))$ is the probability of 
measuring a signal ($y_1$, $y_2$, $y_3$). This is given by taking 
the discrete integral over the collection of reference orientations:
\begin{align}
	p(y_k(\hat{\bf s}))
	=
	\sum_{\alpha=1}^{n_{\rm ref}}
	p(y_1, y_2, y_3, y_4|\hat{\bf n}_\alpha)
	p(\hat{\bf n}_\alpha)
\end{align}

From \eqref{eq:bayes} we obtain the key result, a mass probability 
distribution denoting the probability that our dimer is in 
orientation $\hat{\bf n}_{\alpha}$ given a measured signal ($y_1$, 
$y_2$, $y_3$), \textit{i.e.} an estimated mapping from scattering 
measurement to orientation estimate.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Calculation of error}
\label{sec:divergence}
Using \eqref{eq:bayes} we get a snapshot estimation of the instantaneous
orientation of the trapped dimer based on a single measurement from a 
collection of detector signals. Of course in reality these detectors 
will be returning a discrete time series representative of the dimer's
trajectory within the optical trap.

Using the simulation code from \cite{Vigilante2020} we can replicate 
this trajectory and furthermore use \textit{mstm}to calculate the far field 
intensity produced by said dimer. At each time step we measure the signal
produced at each measurement angle and use our neural network to calculate
$p(y_k(\hat{\bf s})|\hat{\bf n}_\alpha, t)$.

Evaluating how well Eq.~\eqref{eq:bayes} performs in predicting $\hat{\bf n}$
is difficult because we have discretised the orientation space meaning 
there will always be some degree of inaccuracy in our prediction. The Kullback-Liebler divergence method provides a measure of the relative
entropy change between two probability distributions:
\begin{align}
	D_{KL}(P(t)\parallel Q(t)) = \sum P(t) \log_2\left[\frac{P(t)}{Q(t)}\right] 
	\label{eq:D_KL}
\end{align}

Where $P(t)$ and $Q(t)$ are two probability distributions we are
comparing, often it is common to use a base 2 log as this converts 
the entropy to base 2. By relative entropy we are talking about 
the useful information gained/lost by switching from $P(t)$ to 
$Q(t)$. In our case $Q(t)$ is simply the distribution given by
Eq.~\eqref{eq:bayes} and $P(t)$ is an idealised distribution, 
one where the model is accurate with 100\% certainty.
\begin{align}
	\label{eq:best}
	p_{best} = P(t) =
	\begin{cases}
		1 & \text{when $\hat{\bf n}_\alpha$ = $\hat{\bf n}_{best}$}\\
		0 & \text{when $\hat{\bf n}_\alpha$ $\neq$ $\hat{\bf n}_{best}$}
	\end{cases}
\end{align}

In reality the distribution from Eq.~\eqref{eq:bayes} will assign 
some non-zero probability to every reference orientation. Therefore 
the relative entropy for a single measurement is given as:
\begin{align}
	D_{KL}(p_{best}\parallel p(\hat{\bf n}_\alpha| y_k(\hat{\bf s}), t))
	= \log_2 \left[\frac{1}{p(\hat{\bf n}_\alpha| y_k(\hat{\bf s}), t)}
	\right]
	\label{eq:kullback}
\end{align}

The relative entropy is a measure if how spread out our distribution
is. With larger values of $D_{KL}$ indicating that our prediction is 
spread out so the neural net is unsure about $\hat{\bf n}_{best}$. 
Whereas a low value of $D_{KL}$ means that our model is more confident 
in $\hat{\bf n}_{best}$. The only draw back being that the relative 
entropy does nothing to indicate accuracy, only how closely our 
distribution matches our ideal result. This will be better 
exemplified later in \ref{sec:test}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Brownian Simulation}
\label{sec:brownian}

We use the Brownian OT package developed by Fung~\textit{et~al} 
\cite{Vigilante2020} to simulate the motion of an asymmetric dimer (Figure~\ref{fig:dimer}a) within an optical trap. We simulate the 
motion of a dimer trapped in a highly focused Gaussian beam by 
calculating the optical forces imparted by the laser, and the 
Brownian force due to the surrounding fluid. We simulated a 
polystyrene dimer ($n = 1.59$) in a suspension of water ($n_{med} 
= 1.33$). As discussed in chapter 4 dimers can be stably trapped
in off-axis orientations. We simulate a dimer (of size ratio 2) that
is trapped in this off-axis orientation by a $5 mW$ laser, due to 
the meta-stability of the trap the dimer is only trapped for second
before it escapes the trap. The dimer's position and orientation 
is shown below:
\begin{figure}[h!]
	\centering
	\begin{subfigure}{0.45\textwidth}
		\subcaption{}
		\includegraphics[width =\textwidth]{fig7a.png}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{fig7b.png}
	\end{subfigure}
	\caption{Simulation results of: (a) the dimer's orientation vector with time, (b) the dimer's [x,y,z] position with time.}
	\label{fig:motion}
\end{figure}

At the same we use \textit{mstm} to replicate the intensity of light
incident on the surfaces of detectors placed around the focus of the
optical trap. The exact number of detectors was initially assumed to 
be arbitrary, in that it made no difference to our estimate whether 
we used 2 angles or 200.

\subsection{Choice on the number of detectors}
When all of the detectors lie in the same plane the expected signal 
can appear identical despite the dimer being in completely different orientations. It should be noted that these pairs are reflected in 
one or more axis which suggests that these are due to the arrangement 
of our detectors. More specifically, if the detectors are placed say 
in the x-y plane then only when the dimer is pointed nearly fully 
upright will the expected signal be entirely unique. This is not an 
issue with typical analytical static light scattering experiments 
because the signal being collected is averaged over a population of 
particles all in different orientations. Whereas in our set up we are concerned with the scattering of a single particle whose scattering is orientationally dependant.

To remedy this we raise the third detector out of the x-y plane; 
as such the expected signals from each reference orientation is 
unique, though with limited resolution. By adding a $4^{th}$ 
detector we can differentiate signals more reliably, improving the 
neural networks performance. In line with our goal of making this 
method viable in a laboratory setting we decided not to increase 
the number of detectors further than 4.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Testing the Neural Network}
\label{sec:test}
We applied Eq.~\eqref{eq:bayes}, taking the reference orientation 
with the highest probability  as our estimate of the dimer's 
instantaneous orientation $\hat{\bf n}_{est}$. To visualise the 
model's performance we plotted the orientation components of our 
estimation $\hat{\bf n}_{\alpha}$ and the dimer's \emph{actual} 
instantaneous orientation $\hat{\bf s}$ versus time. For comparison, 
we also plotted the orientation components of the closest reference orientation, $\hat{\bf n}_{best}$. Assuming a uniform prior of the 
reference orientations $p(\hat{\bf n}_\alpha)$ the neural network's predictions ($\hat{\bf n}_{\alpha}$ from Eq.~\eqref{eq:bayes}) are 
at times reasonable, but there are significant large and random jumps 
away from the correct result (Fig.~\ref{fig:uniform}). 
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{fig5.png}
	\caption{Model's estimation of dimer orientation over the simulation time, assuming uniform prior $p(\hat{\bf n}_\alpha)$, broken up into x, y, and z components for clarity. Blue line denotes the best result we can achieve (the reference orientation $\hat{\bf n}_{best}$ that is closest to the actual orientation), orange line denotes the result provided by eq~\ref{eq:bayes}: where the blue line is not visible, the model's prediction agrees with $\hat{\bf n}_{best}$. Dotted black line is the instantaneous orientation $\hat{\bf s}$.}
	\label{fig:uniform}
\end{figure}

One reason we observe such large jumps in the estimated 
orientation is simply because the neural network is assuming
that the signals are unrelated from one another. We can 
observe this if we take the average relative entropy change
across the entire trajectory:

\begin{align}
	\langle D_{KL}\rangle = \frac{1}{N} \sum_{i=1}^{N} \log_2 \left[\frac{1}{p(\hat{\bf n}_\alpha\parallel y_k(\hat{s}),t_i)}\right]
	\label{eq:score}
\end{align}

Where we get an average entropy change of $\approx 0.604$, in 
base 2 this means that on average the model has narrowed down 
its options from 1 in 30 to 1 in 1.52. This is not reflected
though in fig.~\ref{fig:uniform} which indicates that while 
the model is highly confident in its results it is not at all 
accurate. As demonstrated by fig~\ref{fig:mixing} the 
orientation and signal spaces are uncorrelated with one another 
which results in the neural network to assign orientations 
incorrectly. Combining this fact with use of a uniform prior 
($p(\hat{\bf n}_\alpha)$), there is no constraint on how much 
estimated orientation can change from time-step to time-step.

To improve the estimation we can therefore use knowledge of the 
physical limitations of the object in the trap and its dynamics. 
Imposing a more physically grounded prior, accounting in this case 
for the fact that the rotational motion of the dimer is limited by 
the rotational trap stiffness ($\kappa_r$). Here the prior of the 
reference orientations $p(\hat{\bf n}_\alpha)$ was redefined at 
each time step as a Boltzmann distribution of the physical distance 
between the previous estimate $\hat{\bf n}_{est}(t-\Delta t)$ and 
each reference orientation $\hat{\bf n}_\alpha$. Put simply, we are 
reweighing our estimation based on the size of rotation, with smaller movements being favoured over large movements:
\begin{align}
	p(\hat{\bf n}_\alpha)
	&=\frac{e^{\hat{\bf n}_\alpha 
			\cdot \hat{\bf n}_{est}(t-\Delta t)}}
	{\sum_{\alpha=1}^{n_{\rm ref}}
		e^{\hat{\bf n}_\alpha 
			\cdot \hat{\bf n}_{est}(t-\Delta t)}}
	\label{eq:boltz}
\end{align}

>>>>

\chapter{Simulated Quadrant Photodiode}
\label{chapter:simulated_QPD}

\section{Overview}
In this chapter, we begin with a brief discussion of the working 
principle behind a position detection device. We then provide a 
framework for replicating the scattering signal recorded by a 
quadrant photo diode. We show that extending the framework to 
capture information from an arbitrary shaped particle is simple 
in principle but difficult to interpret, we demonstrate as such 
by comparing the expected signal from a trapped sphere, to that 
of a trapped dimer.

\section{Introduction}

Beyond merely simulating the forces experienced by a spherical 
aggregate we also developed a simulative quadrant photo diode 
to replicate the results from a typical calibration test. This 
builds upon the work from \cite{Rohrbach2002} which applied 
Lorenz-Mie theory to replicate the response signal of a QPD 
being used in back focal-plane interferometry. Rather than be 
constrained to individual spheres we can now consider the 
expected response from any type of spherical aggregate. The
goal of such simulation software would be to simulate the 
expected experimental results from a complex particle whose 
exact shape and size may be difficult to determine.

In order to simulate a typical experimental set up with a QPD 
installed as a position detection system we need to evaluate 
the total magnitude of the electric field incident on the 
photo-diode surface. While trapping a micro-particle, the 
scattered and incident fields combine together and interfere 
with one another. These fields are collected by a condenser 
lens in the far field limit and are focused onto the QPD 
surface, the total intensity can be evaluated as:
\begin{align}
	I(x,y) = \epsilon_0c\left|
	\begin{bmatrix} 
		E_{i,x}(x,y)+E_{s,x}(x,y) \\ 
		E_{i,y}(x,y)+E_{s,y}(x,y) \\ 
		E_{i,z}(x,y)+E_{s,z}(x,y)
	\end{bmatrix} \right|^2 \times step(\theta_{NA}-\theta(x,y))
\end{align}

The last term is simply a representative step term that defines 
the outer limit by which we evaluate the electric field, this is 
analogous to our condenser lens removing noise from other light 
sources by only accepting light at a specific acceptance angle 
defined by its numerical aperture $NA_c$. Depending on the 
relative size of our particle we can adjust the acceptance angle, 
this has very little effect on the transverse signals, but for 
axial evaluations of a trapped particle the numerical aperture 
should be tuned so that the resultant response curve has negative 
slope in order to allow for axial position detection, the method 
for finding this angle $\theta_\Theta$ is discussed in 
\cite{Friedrich2012}.

The incident beam is simple enough to define given our set up 
parameters, for the sake of simplicity we assume that our beam is a Laguerre-Gaussian 
beam of mode $[0.0, 0.0]$ (which is simply a pure Gaussian beam). 
\textit{Ott} uses a point matching approach to approximate the beam shape 
coefficients of the incident field by fitting it to the far field estimate. 
From the QPD's perspective it is receiving light both from the incident 
and scattered beam simultaneously, as such both fields must be expressed
using outgoing vector spherical harmonics.
\begin{align}
	E_{\rm inc}(r)&=
	E_0\sum^\infty_{n}\sum^n_{m=-n}\left[a_{mn}{\bf M}^{(2)}_{nm}({\bf r})+
	b_{nm}{\bf N}^{(2)}_{nm}({\bf r}) \right] 
	\\
	E_{\rm scat}(r)&=E_0 \sum^\infty_n \sum^n_{m=-n} \left[
	p_{nm}{\bf N}_{nm}^{(2)}({\bf r})+q_{nm}{\bf M}_{nm}^{(2)}({\bf r})
	\right] 
\end{align}

Where the superscript (2) denotes an outgoing spherical harmonic
function. In order to compute the scattering from the target particle \textit{ott} uses the $T$-matrix method, this is not essential 
for a simple sphere but is essential for complex shaped particles 
such as dimers. The scattered and incident fields are then combined 
together in the far field to get $I(x,y)$, the quadrant and overall 
signals are calculated via:
\begin{align}
	Q_i &= \sum_{n,m} I(x_{i,n}, y_{i,m}) \\
	S_{x} &= \frac{(Q_1+Q_2)-(Q_3+Q_4)}{\sum I_0(x,y)} \\
	S_{y} &= \frac{(Q_1+Q_3)-(Q_2+Q_4)}{\sum I_0(x,y)} \\
	S_{z} &= \frac{(Q_1+Q_2+Q_3+Q_4}{\sum I_0(x,y)}
\end{align}

Where the denominator is the total intensity on the QPD while there is 
no particle within the trap. You would expect that $S_x$ and $S_y$ are
near identical for equal particle displacements, however due to the 
polarisation of the beam there will be a slight bias to the signals 
for motion along the direction of the polarisation vector. This is 
generally not an issue if the particle trajectory is sampled for a 
long enough time period. By converting from signal units to length
units (see \eqref{eq:conversion_factor}) the trap shape can be 
discerned from the QPD signals, assuming the fluid properties are
known to a high degree of accuracy.

While translational motion has no knock on effects to the QPD signal, 
rotational changes are double counted by both the incident and total 
field. This results in rotational motion being biased in the QPD signal 
- even when collecting signals from isotropic scatterers. To prevent 
this we need to rotate the total field via the inverse rotation matrix
of the dimer.

To confirm that our method is producing accurate results, we ran a comparison
between our simulative QPD and the results from \cite{Rohrbach2002}. Where a 
$300\ nm$ diameter sphere is scanned across the path of a focused Gaussian beam
($\lambda=1064\ nm$, $NA=1.2$), the sphere has a refractive index of 1.57 and is
suspended in water ($n_{med}=1.33$) and the condenser lens has its numerical aperture
set to 0.5 ($\theta_{max} = 30^\circ$). Scanning across all three primary axis 
produced the following response curve:
\begin{figure}[h!]
	\begin{subfigure}{0.475 \linewidth}
		\subcaption{}
		\includegraphics[width=\linewidth]{QPD_axial_tests.png}
	\end{subfigure}
	\begin{subfigure}{0.475 \linewidth}
		\subcaption{}
		\includegraphics[width=\linewidth]{QPD_lat_tests.png}
	\end{subfigure}
	\caption{Comparison between QPD response signal versus work conducted by 
		Rohrbach, single sphere ($r = 150\ nm$, $n=1.57$) is scanned by a $1064\ nm$ laser and the QPD signal recorded. Solid lines represent the signal 
		produced by QPD using \textit{ott} and points represent the signal 
		response collected from \cite{Rohrbach2002}.}
	\label{fig:Rohrbach}
\end{figure}

The discrepancy between our simulated QPD and the results 
from Rohrbach can be attributed to the fact that the position 
to signal error grows as you move further from trap focus. 
As the particle moves further from the trap focus the error 
in the attributed signal grows substantially \cite{Rohrbach2002}.
In most optical trapping experiments we usually are calibrating 
a strong trap where the mean square displacement falls well within 
the linear regime shown in fig.~\ref{fig:Rohrbach}. In which case 
the error between Rohrbach's model and our own is inconsequential 
for calibration purposes. With this any trajectory can be collected 
from the QPD by displacing and rotating the beam accordingly, 
later in chapters \ref{chapter:langevin_dynamics} and \ref{chapter:simulated_detection} we discuss the limits of 
accuracy that can be achieved using back focal plane interferometry 
to characterise non-spherical targets.

\subsection{Position detection}
\label{sec:position_detection}
In order to accurately capture the dynamics of a trapped particle, a 
position detection system is required. There are 3 possible methods 
of position detection: video-analysis, lateral-effect position sensing, 
and photodiodes.

Video analysis is ideally suited for multiple traps or situations where 
precision is not the top priority. Whereas lateral-effect and photodiode 
position detection are two examples of back-focal plane interferometry, 
where the interference pattern produced by the target particle is 
extrapolated to determine its position. In order for video analysis to 
match the accuracy of back-focal plane interferometry methods requires 
the camera's frame rate to exceed $1\ kHz$ which can be difficult to 
achieve while maintaining a decent resolution \cite{Gibson2008}. In 
comparison off the shelf back-focal plane detectors can achieve temporal 
resolutions anywhere from $10-100\ kHz$ \cite{BergSoerensen2004}.

A lateral-effect sensor has a similar output but works using a the 
entire sensor as a single cell analogous to the focal plane of the 
trapping beam. The four corners of the sensor act as anodes connected 
to a base plate cathode, as the beam moves across the surface of the 
detector each anode will experience a different photocurrent depending 
on how close the centre of the interference pattern is to each anode. 
The advantage of a lateral effect detector is that the linear regime 
is much larger than a QPD making it much better for monitoring the 
position of a trapped particle. However, Lateral-effect sensors are 
often limited in their spacial resolution due to high signal-to-noise 
ratios, requiring a high intensity of light on the sensor in order to 
get a clean signal. As a result, most optical force measurements are 
conducted using a QPD as opposed to a lateral-effect sensor, as often 
the displacement is small enough that the signal-displacement curve 
can be considered linear.
\begin{figure}[h!]
	\centering
	\includegraphics[height=\linewidth, angle=270]{QPD_Lateral_effect.pdf}
	\caption{Comparison between QPD and Lateral effect photodiodes.
		The four quadrants of a QPD (left) experience different photocurrents
		based on the total intensity of light incident on each section 
		(labelled A, B, C, D). 
		Whereas a Lateral effect sensor (right) uses the resistive properties
		of the photodiode surface to vary the create different photocurrents 
		passing through the anodes A, B, C, and D.}
\end{figure}

A quadrant photo diode (QPD) is a frequently used position detection 
system for optical tweezers due to their high sampling rate, high 
degree of precision, and ease of set up. The QPD is constructed of 
four photo diodes assembled in a quadrant formation, when a particle 
is trapped the interference pattern produced is focused onto the QPD, 
with the maximum intensity mapping to the particle's centre of mass. 
By summing the voltages of the horizontal and vertical quadrants together 
the particle's centre of mass is tracked in the $x-y$ plane. Axial 
displacement can be estimated by observing the change in the total 
voltage of the QPD. The outputted signal gives an indication of the 
particle's relative displacement from the beam focus, but in order to 
convert the signal to distance units the trap needs to be calibrated 
(assuming a linear response curve).

@@

At just a cursory glance it would appear that there is a clear
relationship between the QPD signal and the orientation of the 
dimer. If we look at the QPD signal produced by a dimer as it is 
rotated we can see that it is equally sensitive to rotations in
either the X-Z or Y-Z plane but there is no change in the signal 
when rotated by in the X-Y plane. 
\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{rotation_figure.png}
	\caption{QPD signals detected by a dimer being rotated while
		centred at the focus of the optical trap. Top right:
		rotation in the X-Z plane. Top left: rotation in the
		Y-Z plane. Bottom: Rotation in the X-Y plane.}
	\label{fig:QPD_rotation}
\end{figure}

@@
